---
title: "BDS: Survival analysis with applications in medicine"
author:
- Mark Clements
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

# Exercise 5. Delta method for variance estimation with a Poisson regression model.

In outline, we will investigate **cause-specific mortality** after a diagnosis of colon cancer. We will seek to estimate the hazards and survival by calendar year of diagnosis, including standard errors and 95% confidence intervals.

```{r, message=FALSE}
library(biostat3) # colon
library(dplyr)
```

## (a) Split by time

Using the `biostat3::colon` data-frame, split time from diagnosis (`surv_mm`) by month. Use either `survSplit` or `dplyr`. Check that the total person-time and number of events is the same before and after splitting.

```{r}
split_data <- colon |>
  transform(
    event = status == "Dead: cancer",
    x = 0 + (year8594 == "Diagnosed 85-94")
  ) |>
  survSplit(
    Surv(surv_mm, event) ~ .,
    data = _,
    cut = seq(0, max(colon$surv_mm)),
    episode = "timeband"
  ) |>
  select(-subsite, -surv_yy, -year8594, -ydx, -yexit) |>
  transform(
    pt = surv_mm - tstart,
    mid_t = (tstart + surv_mm) / 2
  )

total_pt_org <- sum(colon$surv_mm)
total_pt_split <- sum(split_data$pt)
stopifnot(total_pt_org == total_pt_split)

total_events_org <- sum(colon$status == "Dead: cancer")
total_events_split <- sum(split_data$event)
stopifnot(total_events_org == total_events_split)

with(biostat3::colon, c(
  nrow = length(status),
  total_events = sum(status == "Dead: cancer"),
  total_person_time = sum(surv_mm)
))
with(split_data, c(
  nrow = length(status),
  total_events = sum(event),
  total_person_time = sum(pt)
))
```

## (b) Poisson model fit

Fit a Poisson regression model adjusting for time from diagnosis, stage of diagnosis and age group at diagnosis. You can choose the functional form for modelling time from diagnosis. Write out the regression model and interpret the parameters.

```{r}
model <- glm(
  formula = event ~ surv_mm + stage + agegrp + offset(log(pt)),
  data = split_data
)
summary(model)
```

## (c) Hazard predictions

### (i)

Predict the hazard for a patient aged 55 years diagnosed with distant cancer at 5 and 10 years. 

```{r}
new_data <- data.frame(
  surv_mm = c(5 * 12, 10 * 12),
  stage = factor("Distant", levels = levels(split_data$stage)),
  agegrp = factor("45-59", levels = levels(split_data$agegrp)),
  pt = 1
)
log_hazard <- predict(model, newdata = new_data, type = "link")
hazard <- exp(log_hazard)
data.frame(
  time_years = new_data$surv_mm / 12,
  log_hazard = log_hazard,
  hazard = hazard
)
```

### (ii)

What is the design matrix for the linear predictor for the log hazards? Given the variance-covariance matrix for the regression coefficients and the design matrix, show how to calculate the standard error for the linear predictor for the the log hazards. Moreover, calculate the 95% confidence intervals for the log hazards.

```{r}
# design_matrix <- model.matrix(model)
design_matrix <- model.matrix(~ surv_mm + stage + agegrp, data = new_data)
design_matrix

covariance_matrix <- vcov(model)
covariance_matrix

var_eta <- diag(design_matrix %*% covariance_matrix %*% t(design_matrix))
se_eta <- sqrt(var_eta)

# Compute confidence intervals
ci_lower <- log_hazard - 1.96 * se_eta
ci_upper <- log_hazard + 1.96 * se_eta

# Create a data frame with results
results <- data.frame(
  time_years = new_data$surv_mm / 12,
  log_hazard = log_hazard,
  se_log_hazard = se_eta,
  ci_lower = ci_lower,
  ci_upper = ci_upper
)

print(results)
```

### (iii)

Check your calculations for (ii) using `predict(..., se.fit=TRUE)`.

```{r}
predict(model, newdata = new_data, type = "link", se.fit = TRUE)
```

## (c) Survival predictions

### (i)

We can use `predict(object, newdata, type="response")` to predict the hazard at a given time. Write a function `h` with signature `function(t)` to predict the hazard for a patient aged 55 years diagnosed with distant cancer at time `t`. Ensure that this function allows for `t` to be a vector and return a vector of hazards (e.g. `h(1:10)`).

```{r}
h <- function(t) {
  # Create a new data frame for the given time points
  new_data <- data.frame(
    surv_mm = t * 12, # Convert years to months
    stage = factor("Distant", levels = levels(split_data$stage)),
    agegrp = factor("45-59", levels = levels(split_data$agegrp)),
    pt = 1
  )

  # Predict hazard using type="response" (which exponentiates log-hazard)
  hazard <- predict(model, newdata = new_data, type = "response")

  return(pmax(hazard, 0))
}
```

### (ii) 

Using the function `h` from (i) and the `integrate` function, calculating the cumulative hazard to 60 months and 120 months (hint: use `value` from the `integrate` object). Calculate survival at 60 and 120 months.

```{r}
cumulative_hazard <- function(t) {
  result <- integrate(h, lower = 0, upper = t)
  return(result$value) # Extract the numerical result of the integration
}
H_60 <- cumulative_hazard(5) # 5 years
H_120 <- cumulative_hazard(10) # 10 years

# Compute survival probabilities
S_60 <- exp(-H_60)
S_120 <- exp(-H_120)

# Print results
results <- data.frame(
  time_months = c(60, 120),
  cumulative_hazard = c(H_60, H_120),
  survival_probability = c(S_60, S_120)
)

print(results)
```

### (iii) 

Write a function `grad` to calculate the gradient of some function `g` with respect to a vector `x` using two-sided finite differences. Specifically, 
\begin{align*}
\frac{\partial g(\boldsymbol{x})}{\partial x_j} &= \frac{g(\boldsymbol{x}+\boldsymbol{e}_j \epsilon)-g(\boldsymbol{x}-\boldsymbol{e}_j \epsilon)}{2\epsilon}
\end{align*}
where $\epsilon=10^{-5}$ and $\boldsymbol{e}_j$ is a unit vector which includes one at the $j$-th position and otherwise zeros. 


```{r}
grad <- function(g, x) {
  ## returns the gradient of g(x) with respect to each coefficient
  epsilon <- 1e-5 # Small perturbation value
  n <- length(x) # Number of elements in x
  gradient <- numeric(n) # Initialize gradient vector

  for (j in 1:n) {
    e_j <- rep(0, n) # Create unit vector
    e_j[j] <- epsilon # Set perturbation at the j-th position

    # Compute finite difference approximation
    gradient[j] <- (g(x + e_j) - g(x - e_j)) / (2 * epsilon)
  }

  return(gradient)
}

g <- function(x) sum(x^2)
g(3)
g(3:4)
grad(g, 1:3)
```

### (iv)

Use `grad` to calculate the gradients for predicted survival at 60 and 120 months with respect to the regression model coefficients (hint: your function `g` will need to change the `coefficients` in the `glm` object). Using the delta method, use the gradients and the variance-covariance matrix to calculate the standard errors for survival at 60 and 120 months. Finally, calculate the 95% confidence interval for survival at 60 and 120 months.

```{r}
# Function to compute survival probability for given model coefficients
g <- function(beta, t) {
  # Update the model coefficients
  model$coefficients <- beta

  # Define hazard function h(t) based on updated model
  h <- function(t) {
    new_data <- data.frame(
      surv_mm = t * 12,
      stage = factor("Distant", levels = levels(split_data$stage)),
      agegrp = factor("45-59", levels = levels(split_data$agegrp)),
      pt = 1
    )

    return(predict(model, newdata = new_data, type = "response"))
  }

  # Compute cumulative hazard H(t)
  H_t <- integrate(h, lower = 0, upper = t)$value

  # Compute survival probability S(t)
  S_t <- exp(-H_t)

  return(S_t)
}

# Compute gradient using finite differences
grad_survival_60 <- grad(function(beta) g(beta, 5), model$coefficients)
grad_survival_120 <- grad(function(beta) g(beta, 10), model$coefficients)

# Extract variance-covariance matrix
cov_matrix <- vcov(model)

# Compute variance using the delta method
var_survival_60 <- t(grad_survival_60) %*% cov_matrix %*% grad_survival_60
var_survival_120 <- t(grad_survival_120) %*% cov_matrix %*% grad_survival_120

# Compute standard errors
se_survival_60 <- sqrt(var_survival_60)
se_survival_120 <- sqrt(var_survival_120)

# Compute predicted survival probabilities
survival_60 <- g(model$coefficients, 5)
survival_120 <- g(model$coefficients, 10)

# Compute 95% confidence intervals
ci_lower_60 <- survival_60 - 1.96 * se_survival_60
ci_upper_60 <- survival_60 + 1.96 * se_survival_60

ci_lower_120 <- survival_120 - 1.96 * se_survival_120
ci_upper_120 <- survival_120 + 1.96 * se_survival_120

# Create results data frame
results_survival <- data.frame(
  time_months = c(60, 120),
  survival_estimate = c(survival_60, survival_120),
  se_survival = c(se_survival_60, se_survival_120),
  ci_lower = c(ci_lower_60, ci_lower_120),
  ci_upper = c(ci_upper_120, ci_upper_120)
)

# Print results
print(results_survival)
```
